Reversible Modular Addition using a Dirty Bit

### 2017-03-28 paper.tex
Cleaning

\begin{abstract}
\em
We present size-efficient reversible classical circuits for performing addition, incrementation, comparisons, modular addition, and modular offset on registers of differing sizes using no ancilla or 1 ancilla in an unknown state.
We improve the number of qubits needed to factor an $n$-bit number from $2n+2$ \cite{takahashi2006, haner2016} to $2n+1$, without increasing the asymptotic cost of the circuit.
\end{abstract}

@@

When constructing quantum circuits, or classical reversible circuits, the number of available ancilla is an important resource.
An ancilla is an extra bit or qubit, on top of those used in input and output roles, available for use as temporary workspace.
Ancilla may be initialized to a known state (``clean bit"), or be given to the circuit in an unknown state (``dirty bit") that must be restored before the circuit finishes.
Clean bits are more valuable, allowing for simpler and more compact circuit constructions, but dirty bits are more plentiful, because any temporarily unused qubit is a borrowable dirty bit.

In this paper we focus on circuits that use at most one dirty bit.

Because one part of a circuit can borrow dirty bits from another part of the same circuit, circuit constructions that require only dirty bits are easier to apply under tight space constraints, or on circuit topologies where other ancilla are too far away to be acquired quickly.
For example, H\"{a}ner et al. used dirty bits to improve the cost of performing Shor's algorithm, when limited to using only $2n+2$ qubits and Toffoli based circuits, to $O(n^3)$ depth and $O(n^3 \lg n)$ size \cite{haner2016}.
We improve on \cite{haner2016} by cutting the number of required qubits from $2n+2$ to $2n+1$.

Our paper is structured as follows.
The introduction describes the various sections and also itself for no discernible reason.
Section \ref{sec:review} discusses existing circuit constructions.
In sections \ref{sec:first-circuit} through \ref{sec:last-circuit} we describe various arithmetic circuit constructions that use 0 or 1 dirty bit, building up to modular addition.
We then use our single-dirty-bit modular addition to factor an $n$ bit number using $2n+1$ qubits in section \ref{sec:applications}.

@@

We generalize the ideas in \cite{takahashi2005} to apply to additions where the input register may be smaller than the target register.

@@

An offset gate adds a compile-time constant into a register.
Many reversible adder constructions do not work as offset gates because there is no input register to use as temporary workspace.

Draper's QFT-based addition \cite{draper2000} doesn't need the source register as workspace, and so can be applied to this problem.

\cite{haner2016} describes a divide-and-conquer offset circuit that requires $O(N)$ depth, $O(N \lg N)$ size, and $n/2$ dirty ancilla.
[[[ Actually, only 1 is required by their construction. 0 if you go quantum. And when I emailed them about it they said they would fix the oversight. So I'm not sure how to describe that. It is important for this paper that we have an non-modular offset circuit that only needs 1 dirty ancilla. ]]]

We describe a circuit for the special case of offset-by-1, i.e. incrementation.
Our circuit uses $O(N)$ depth, $O(N)$ size, and a single dirty ancilla.
We also note how to remove the ancilla with quantum operations.
Our construction is a simplified version of our unpublished incrementer described in \cite{gidney2015}.

@@

Comparison operators, which toggle a target bit based on if one input is larger, smaller, and/or equal to another, can be implemented efficiently using adders \cite{takahashi2005}.
Due to generalizing Takahashi's adder \cite{takahashi2005} to apply to different sizes, we also generalize Takahashi's comparators to apply to registers of different sizes.

@@

\cite{vedral1995} describes a modular adder that uses $O(N)$ clean ancilla, based on performing normal addition and subtraction and temporarily storing a comparison in a clean ancilla.
Most of the clean ancilla are only used for the addition and subtraction, so using inline adders reduces the number of clean ancilla required to 1, as used in \cite{haner2016}.

[[[ do better here ]]]

We improve on this construction by allowing the ancilla to be dirty.

@@

The need for the ancilla to be clean can be fixed by inverting (bitwise-NOT-ing) the input and target registers before and after the addition when the ancilla is on.
Because $\lnot x = -x-1$, and the registers are the same size, temporarily inverting the target effectively turns the addition into a subtraction.
Temporarily inverting the input turns the subtraction back into an addition, but also introduces an extra shift by -1, cancelling out the +1 error.

@@

Because this adder construction uses the input register as workspace, it only works when the input register is the same size as the target register.
When the input register is larger the high bits of the input register can simply be ignored, but when the target register is larger the workaround is more difficult.
Not only is there insufficient workspace, but the register sizes being different breaks the logical-negation trick that allowed the ancilla to be dirty.

To work around the target register being larger, three fixes are needed.
Each uses a controlled increment or decrement gate, which will be described in the next section.
(Note that, to avoid cyclic dependencies, the increment and decrement constructions will only use the same-register-size adder.)
First, the carry signal must be forwarded into the high part of the target register with a controlled increment instead of with a CNOT.
Second, to free the MSB of the input register for use as the carry signal, we add it into the target ahead of time with a controlled increment.
Finally, when the MSB is on instead of off its role as carry signal causes an extra increment of the target.
This extra increment is cancelled with a controlled decrement.
See the circuit diagram in figure \ref{fig:inline-adder-into-large} for the complete construction.

@@

To improve from $N$ dirty bits to a single dirty bit, we break the increment operation into two halves.
A top half that is incremented only if all of the bottom bits are on, and a bottom half that is unconditionally incremented.
The bottom half can be incremented with the double-subtraction trick by borrowing the top half.
But the double-subtraction trick doesn't work on the top half, because the bottom half can't be borrowed when it is being used as a control.

To work around not being able to operate on the borrowed bits while using them as a control, start with the double-subtraction trick but change the second subtraction to an addition and conditionally toggle the target bits instead of the input bits.
When the condition isn't satisfied, the addition and subtraction will cancel each other.
When the condition is satisfied, the addition will be inverted into a subtraction, and both the subtraction and addition-turned-subtraction will fire, subtracting the input register from the target register twice.
However, because we are conditioning on the input bits all being on, we know the input must be -1.
Therefore the target was incremented by 2.
To halve the +2 into a +1, prepend a dirty LSB onto the target register as shown in \ref{fig:compact-increment}.

@@

Because the adder/subtracter construction used by the incrementer in figure \ref{fig:compact-increment} requires the target and source registers to be the same size, the construction described so far only works on odd-sized registers.
For even-sized registers, another minor variation on the conditionally-invert-addition-into-subtraction-with-logical-negation technique is used.
That construction is shown in figure \ref{fig:compact-increment-even}.

@@

As with addition and subtraction, you can turn an increment into a decrement either by reversing the gate order, by bordering the circuit with NOT gates, or by inverting all of the controls.
Also, increment gates can be efficiently controlled by using another invert-input-and-output trick (see figures \ref{fig:controlled-increment-odd} and \ref{fig:controlled-increment-even}).

@@

Beware that, when the $B$ register is smaller than the $A$ register, the high bits of $A$ are not accounted for by the subtraction and addition and so the construction breaks.
To fix the problem, swap the roles of A and B (i.e. subtract out of A instead) and use the opposite comparison.
Alternatively, any operations generated by the subtraction or decrement gates that toggle the target bit can be controlled by all of the high bits of $A$ being off (and then, if any of those high bits are on, meaning $A$ must be larger than $B$, toggle the result bit as appropriate for the comparison).

@@

In \cite{haner2016}, H\"{a}ner et al. describe how to perform Shor's algorithm with Toffoli-based circuits using $2n+2$ qubits.
The first $n$ qubits are needed to hold the input.
The second $n$ qubits are for holding the output of exponentiating the input modulo a constant.
One of the remaining qubits is used in the phase estimation process.
The last qubit is a clean ancilla used by their modular offset operation to temporarily hold a comparison against the modulus.

The modular addition circuit we have presented does not require that last clean ancilla.
Instead, we borrow the phase estimation qubit as a temporary dirty bit.
Our modular offset circuit is a drop-in replacement for the one used by H\"{a}ner et all.
It even has the same asymptotic costs: $O(N \lg N)$ size and $O(N)$ depth.
Using our modular offset circuit, and borrowing the phase estimation qubit instead of using a separate clean bit, creates an implementation of Shor's factoring algorithm that uses only $2n+1$ qubits without increasing the size or depth complexity.

@@

We present classical reversible circuits for incrementing, modular addition, and modular offset that use only a single dirty bit.
Our modular offset circuit improves the minimum number of qubits used when factoring an $n$-bit integer from $2n+2$ to $2n+1$.

>>>>

\begin{abstract}
\em
We present size-efficient reversible classical circuits for performing addition, incrementation, comparisons, modular addition, and modular offset on registers of differing sizes using either no ancilla or 1 ancilla in an unknown state.
We improve the number of qubits needed to factor an $n$-bit number from $2n+2$ \cite{takahashi2006, haner2016} to $2n+1$, without increasing the asymptotic cost of the circuit.
\end{abstract}

@@

When constructing quantum circuits, or classical reversible circuits, the number of available ancilla is an important resource.
An ancilla is an extra bit or qubit available for use by a circuit as temporary workspace.
Ancilla may be initialized to a known state (``clean bit"), or be given to the circuit in an unknown state (``dirty bit") that must be restored before the circuit finishes.
Clean bits are more valuable, allowing for simpler and more compact circuit constructions, but dirty bits are more plentiful, because any temporarily unused qubit is a borrowable dirty bit.

Because one part of a circuit can borrow dirty bits from another part of the same circuit, circuit constructions that require only dirty bits are easier to apply under tight space constraints, or on circuit topologies where other ancilla are too far away to be acquired quickly.
For example, H\"{a}ner et al. used dirty bits to improve the cost of performing Shor's algorithm, when limited to using only $2n+2$ qubits and Toffoli based circuits, to $O(n^3)$ depth and $O(n^3 \lg n)$ size \cite{haner2016}.
The modular addition construction in this paper improves on \cite{haner2016} by cutting the number of required qubits to factor an $n$-bit number from $2n+2$ to $2n+1$.

Our paper is structured as follows.
The introduction describes the various sections and also itself for no discernible reason.
Section \ref{sec:review} discusses existing circuit constructions.
In sections \ref{sec:first-circuit} through \ref{sec:last-circuit} we describe various arithmetic circuit constructions that use 0 or 1 dirty bit, building up to modular addition.
We then describe how to use our single-dirty-bit modular addition to factor an $n$ bit number using $2n+1$ qubits in section \ref{sec:applications}.

@@

We generalize the ideas in \cite{takahashi2005} to apply to additions where the input and target registers have differing sizes.

@@

An offset gate adds a compile-time constant into a register.
Some reversible adder constructions can be used as offset gates, but others can't because they use qubits in the input register as temporary workspace.
An offset gate has no input register.

Draper's QFT-based addition construction \cite{draper2000} doesn't use the input register as workspace, and so trivially generalizes into an offset gate.

\cite{haner2016} describes a Toffoli-based divide-and-conquer offset circuit that requires $O(N)$ depth, $O(N \lg N)$ size, and $n/2$ dirty ancilla.
[[[ Actually, only 1 is required by their construction. 0 if you go quantum. And when I emailed them about it they said they would fix the oversight. So I'm not sure how to describe that. It is important for this paper that we have an non-modular offset circuit that only needs 1 dirty ancilla. ]]]

We describe a circuit for the special case of offset-by-1, i.e. incrementation.
Our increment circuit uses $O(N)$ depth, $O(N)$ size, one dirty ancilla, and no non-classical operations.
We also note how to cut the ancilla with quantum operations.
This construction is a simplified version of our unpublished incrementer described in \cite{gidney2015}.

@@

Comparison operators, which toggle a target bit based on if one input is larger, smaller, and/or equal to another, can be implemented efficiently using adders \cite{takahashi2005}.

Due to generalizing Takahashi's adder \cite{takahashi2005} to apply to different sizes, we also generalize Takahashi's comparators to apply to registers of different sizes.

@@

\cite{vedral1995} describes a modular adder that uses $O(N)$ clean ancilla.
It temporarily stores a comparison in a clean ancilla, performs a normal addition, uses a subtraction conditioned on the ancilla to wrap the result back into range, then performs a cleanup comparison to clear the ancilla.

All but one of the $O(N)$ clean ancilla used by \cite{vedral1995} are due to using inefficient addition circuits.
Using later zero-ancilla adder constructions reduces xthe number of clean ancilla required to perform a modular addition to just one.

Our modular adder construction also requires a single ancilla, but the ancilla is no longer required to be clean.

@@

The need for the ancilla to be clean can be fixed by inverting (bitwise-NOT-ing) the input and target registers before and after the addition when the ancilla is on.
Because $\lnot x = -x-1$, and the registers are the same size, temporarily inverting the target effectively turns the addition into a subtraction.
Temporarily inverting the input effectively turns the subtraction back into an addition, but also introduces an extra shift by -1, cancelling out the +1 error.

@@

Because this adder construction uses the input register as workspace, it only works when the input register is the same size as the target register.
When the input register is larger, the high bits of the input register can simply be ignored.
But, when the target register is larger, the workaround is more difficult.
Not only is there insufficient workspace, but the register sizes being different breaks the logical-negation trick that allowed the ancilla to be dirty.

To work around the target register being larger, three fixes are needed.
Each uses a controlled increment or decrement gate, which will be described in the next section.
(Note that, to avoid cyclic dependencies, the increment and decrement constructions will only use the same-register-size adder.)

First, the carry signal must be forwarded into the high part of the target register with a controlled increment instead of with a CNOT.
Second, to free the MSB of the input register for use as the carry signal, we add it into the target ahead of time with a controlled increment.
Finally, when the MSB is on instead of off its role as carry signal causes an extra increment of the target.
This extra increment is cancelled with a controlled decrement.
See the circuit diagram in figure \ref{fig:inline-adder-into-large}.

@@

To improve from $N$ dirty bits to a single dirty bit, we break the increment operation into two halves.
A high half that is incremented only if all of the bottom bits are on, and a low half that is unconditionally incremented.
The low half can be incremented with the double-subtraction trick by borrowing the high half.
But the double-subtraction trick doesn't work on the high half, because the low half can't be borrowed when it is being used as a control.

To work around not being able to operate on the borrowed bits while using them as a control, start with the double-subtraction trick but change the second subtraction to an addition and conditionally toggle the target bits instead of the input bits.
When the condition isn't satisfied, the addition and subtraction will cancel each other.
When the condition is satisfied, the addition will be inverted into a subtraction, and both the subtraction and addition-turned-subtraction will fire, subtracting the input register from the target register twice.
However, because we are conditioning on the input bits all being on, we know the input must be -1.
Therefore the target was incremented by 2.
To halve the +2 into a +1, prepend a dirty least-significant-bit onto the target register as shown in \ref{fig:compact-increment}.

@@

Because the adder/subtracter construction used by the incrementer in figure \ref{fig:compact-increment} requires the target and source registers to be the same size, the incrementer construction described so far only works on odd-sized registers.
For even-sized registers, another minor variation on the conditionally-invert-addition-into-subtraction-with-logical-negation technique is used.
That construction is shown in figure \ref{fig:compact-increment-even}.

@@

As with addition and subtraction, you can turn an increment into a decrement either by reversing the gate order, by bordering the circuit with NOT gates, or by inverting controls.
Also, increment gates can be efficiently controlled by using another invert-input-and-output trick (see figures \ref{fig:controlled-increment-odd} and \ref{fig:controlled-increment-even}).

@@

Beware that, when the $B$ register is smaller than the $A$ register, the high bits of $A$ are not accounted for by the subtraction and addition and so the construction breaks.
To fix the problem, swap the roles of A and B (i.e. subtract out of A instead) and use the opposite comparison.
Alternatively, any operations generated by the subtraction or decrement gates that toggle the target bit can be controlled by all of the high bits of $A$ being off and otherwise (if any of the high bits are on) concluding $A>B$ and toggling the result bit as appropriate for the comparison being implemented.

@@

In \cite{haner2016}, H\"{a}ner et al. describe how to perform Shor's algorithm with Toffoli-based circuits using $2n+2$ qubits.
The first $n$ qubits are needed to hold the input.
The second $n$ qubits are used for performing a non-power-of-2 QFT [[[ cite what they used ]]].
One qubit is used for the phase estimation process [[[ cite what they used ]]].
The last qubit is a clean ancilla used by their modular offset operation (to temporarily hold a comparison result).

The modular addition circuit we have presented does not require that last clean ancilla.
Instead, we borrow the phase estimation qubit as a temporary dirty bit.
Our modular offset circuit is a drop-in replacement for the one used by H\"{a}ner et all, with the same asymptotic costs ($O(N \lg N)$ size and $O(N)$ depth).
Doing so creates an implementation of Shor's factoring algorithm that uses $2n+1$ qubits, without increasing the asymptotic size or depth with respect to the $2n+2$ qubit algorithm.

@@

We present linear sized classical reversible circuits for incrementing, modular addition, and modular offset that use only a single dirty bit.
Our modular offset circuit improves the minimum number of qubits used when factoring an $n$-bit integer from $2n+2$ to $2n+1$.
